Bandit Multiclass Linear Classification: \\ Efficient Algorithms for the Separable Case

### 2019-05-03 ICML-camera-ready/empirical-comparison.tex
add camera-ready folder

\begin{figure}[h]
  \centering
\begin{subfigure}[b]{0.3\textwidth}
\captionsetup{justification=centering}
\begin{center}
\includegraphics[width=\textwidth, trim={0, 0cm, 0, 0}, clip]{figures/strong_points.pdf}
\caption{Strongly separable case}
\label{figure:strongly-separable-dataset}
\end{center}
\end{subfigure}
\begin{subfigure}[b]{0.3\textwidth}
\captionsetup{justification=centering}
\centering
\includegraphics[width=\textwidth, trim={0, 0cm, 0, 0}, clip]{figures/weak_points.pdf}
\caption{Weakly separable case}
\label{figure:weakly-separable-dataset}
\end{subfigure}
\caption{Strongly and weakly linearly separable datasets in $\R^3$ with $K=3$
classes and $T=5\times 10^6$ examples. Here we show projections of the examples
onto their first two coordinates, which lie in the ball of radius $1/\sqrt{2}$
centered at the origin. The third coordinate is $1/\sqrt{2}$ for all examples.
Class 1 is depicted red. Classes 2 and 3 are depicted green and blue,
respectively. $80\%$ of the examples belong to class 1, $10\%$ belong to class 2
and $10\%$ belong to class 3. Class 1 lies in the angle interval $[-15^\circ,
15^\circ]$, while classes 2 and 3 lie in the angle intervals $[15^\circ,
180^\circ]$ and $[-180^\circ, -15^\circ]$ respectively. The examples are
strongly and weakly linearly separable with a margin of $\gamma=0.05$,
respectively. (Examples lying within margin $\gamma$ of the linear separators
were rejected during sampling.)}
\label{figure:strongly-and-weakly-separable-datasets}
\end{figure}

>>>>

\begin{figure}[h]
  \centering
\begin{subfigure}[b]{0.3\textwidth}
\captionsetup{justification=centering}
\begin{center}
\includegraphics[width=\textwidth, trim={0, 0cm, 0, 0}, clip]{figures/strong_points}
\caption{Strongly separable case}
\label{figure:strongly-separable-dataset}
\end{center}
\end{subfigure}
\begin{subfigure}[b]{0.3\textwidth}
\captionsetup{justification=centering}
\centering
\includegraphics[width=\textwidth, trim={0, 0cm, 0, 0}, clip]{figures/weak_points}
\caption{Weakly separable case}
\label{figure:weakly-separable-dataset}
\end{subfigure}
\caption{Strongly and weakly linearly separable datasets in $\R^3$ with $K=3$
classes and $T=5\times 10^6$ examples. Here we show projections of the examples
onto their first two coordinates, which lie in the ball of radius $1/\sqrt{2}$
centered at the origin. The third coordinate is $1/\sqrt{2}$ for all examples.
Class 1 is depicted red. Classes 2 and 3 are depicted green and blue,
respectively. $80\%$ of the examples belong to class 1, $10\%$ belong to class 2
and $10\%$ belong to class 3. Class 1 lies in the angle interval $[-15^\circ,
15^\circ]$, while classes 2 and 3 lie in the angle intervals $[15^\circ,
180^\circ]$ and $[-180^\circ, -15^\circ]$ respectively. The examples are
strongly and weakly linearly separable with a margin of $\gamma=0.05$,
respectively. (Examples lying within margin $\gamma$ of the linear separators
were rejected during sampling.)}
\label{figure:strongly-and-weakly-separable-datasets}
\end{figure}

### 2019-05-03 ICML-camera-ready/mistake-lower-bound-for-ignorant-algorithms.tex
add camera-ready folder

In rounds $1,2,\dots,\tau-1$, the algorithm predicts the incorrect class $2$
with probability at least $q_0$. Thus,
$$
\Exp\left[\sum_{t=1}^{\tau - 1} \indicator{\widehat y_t\neq y_t}\right] = q_0 \Exp[(\tau - 1)] \; .
$$
In rounds $\tau, \tau+1, \dots, T$, all the examples are the same and are equal
to $(v_\tau, 2)$. Let $s$ be the first time step $t$ such that $t \ge \tau$
and the algorithm makes a correct prediction. If the algorithm makes mistakes
in all rounds $\tau, \tau+1, \dots, T$, we define $s = T+1$.
By definition the algorithm makes mistakes in rounds $\tau, \tau+1, \dots, s-1$.
Therefore,
$$
\Exp\left[\sum_{t=\tau}^T \indicator{\widehat y_t\neq y_t}\right] \ge \Exp[s-\tau].
$$
%Since the algorithm is ignorant, conditioned on $\tau$, $s-\tau+1$ has the same distribution as
%$\min(X, T-\tau+2)$, where $X$ is drawn from a geometric distribution with parameter
%$p_\tau(2|v_\tau)$.
Since the algorithm is ignorant, conditiond on $\tau$, $s-\tau$ follows a truncated 
geometric distribution with parameter $q\triangleq p_\tau(2|v_\tau)$
(i.e., $s-\tau$ is $0$ with probability $q$, $1$ with probability $(1-q)q$, $2$ with probability $(1-q)^2q, \ldots$). Its expectation can be calculated as follows: 
\begin{align*}
\Exp[s-\tau~|~\tau] 
&= \sum_{i=1}^{T+1-\tau} i\times  \Pr[s-\tau=i]= \sum_{i=1}^{T+1-\tau}\sum_{j=1}^i \Pr[s-\tau=i] = \sum_{j=1}^{T+1-\tau}\sum_{i=j}^{T+1-\tau} \Pr[s-\tau=i] \\
&= \sum_{j=1}^{T+1-\tau}\Pr[s-\tau\geq j] = \sum_{j=1}^{T+1-\tau}(1-q)^j \geq  \sum_{j=1}^{T+1-\tau}(1-q_0)^j = \frac{1-q_0}{q_0}\left( 1-(1-q_0)^{T-\tau+1} \right). 
\end{align*}
%By Lemma~\ref{lem:truncated-geom} below,
%\[
%  \Exp\left[s-\tau+1 \mid \tau, p_\tau(2|v_\tau) \right] = F(p_\tau(2|v_\tau), T-\tau+2) \geq F(q_0, T-\tau+2).
%\]
%Therefore,
%$$
%\Exp[s-\tau ~|~ \tau]
%\ge
%F(q_0, T-\tau+2) - 1
%=
%\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right) \; .
%$$

@@

Combining the two, we have that
\begin{align*}
 \Exp \left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right]
& \ge q_0 \Exp[\tau - 1] + \Exp \left[\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right)\right] \\
& =  \Exp \left[ \Exp \left[ q_0 (\tau - 1) + \frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right) ~\middle|~ \tau \right] \right] \; .
\end{align*}
We lower bound the last expression by considering two cases for $\tau$.
If $\tau \ge \frac{1}{2}T + 1$, then the last expression is lower bounded by
$\frac{1}{2}q_0 T = \frac{1}{2} \sqrt{T}$. If
$\tau < \frac{1}{2}T+1$, it is lower bounded by
\begin{align*}
& \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2}T}\right) \\
& = \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2q_0^2}}\right) \\
& \ge \frac{1-\frac{1}{\sqrt{2}}}{q_0}\left(1-\frac{1}{\sqrt{e}}\right) \\
& \ge \frac{1}{10} \sqrt{T} \; .
\end{align*}

>>>>

In rounds $1,2,\dots,\tau-1$, the algorithm predicts the incorrect class $2$
with probability at least $q_0$. Thus,
\begin{equation}
\Exp\left[\sum_{t=1}^{\tau - 1} \indicator{\widehat y_t\neq y_t}\right] = q_0 \Exp[(\tau - 1)] \; .
\label{eqn:err-phase1}
\end{equation}
In rounds $\tau, \tau+1, \dots, T$, all the examples are the same and are equal
to $(v_\tau, 2)$. Let $s$ be the first time step $t$ such that $t \ge \tau$
and the algorithm makes a correct prediction. If the algorithm makes mistakes
in all rounds $\tau, \tau+1, \dots, T$, we define $s = T+1$.
By definition the algorithm makes mistakes in rounds $\tau, \tau+1, \dots, s-1$.
Therefore,
\begin{equation}
\Exp\left[\sum_{t=\tau}^T \indicator{\widehat y_t\neq y_t}\right] \ge \Exp[s-\tau].
\label{eqn:err-phase2}
\end{equation}
%Since the algorithm is ignorant, conditioned on $\tau$, $s-\tau+1$ has the same distribution as
%$\min(X, T-\tau+2)$, where $X$ is drawn from a geometric distribution with parameter
%$p_\tau(2|v_\tau)$.
Since the algorithm is ignorant, conditioned on $\tau$ and $q \triangleq p_\tau(2|v_\tau)$, $s-\tau$ follows a truncated geometric distribution with parameter $q$
(i.e., $s-\tau$ is $0$ with probability $q$, $1$ with probability $(1-q)q$, $2$ with probability $(1-q)^2q, \ldots$). Its conditional expectation can be calculated as follows:
\begin{align*}
\Exp[s-\tau~|~\tau, q]
&= \sum_{i=1}^{T+1-\tau} i \times \Pr[s-\tau = i |~\tau, q ] \\
&= \sum_{j=1}^{T+1-\tau}\Pr[s-\tau\geq j | ~\tau, q] = \sum_{j=1}^{T+1-\tau}(1-q)^j \geq  \sum_{j=1}^{T+1-\tau}(1-q_0)^j = \frac{1-q_0}{q_0}\left( 1-(1-q_0)^{T-\tau+1} \right).
\end{align*}

Therefore, by the tower property of conditional expectation,
\[ \Exp[s-\tau~|~\tau] = \Exp\left[ \Exp\left[s-\tau~\middle|~\tau, q\right]~ \middle| ~\tau\right] \geq \frac{1-q_0}{q_0}\left( 1-(1-q_0)^{T-\tau+1}\right). \]

%= \sum_{i=1}^{T+1-\tau}\sum_{j=1}^i \Pr[s-\tau=i] = \sum_{j=1}^{T+1-\tau}\sum_{i=j}^{T+1-\tau} \Pr[s-\tau=i] \\
%By Lemma~\ref{lem:truncated-geom} below,
%\[
%  \Exp\left[s-\tau+1 \mid \tau, p_\tau(2|v_\tau) \right] = F(p_\tau(2|v_\tau), T-\tau+2) \geq F(q_0, T-\tau+2).
%\]
%Therefore,
%$$
%\Exp[s-\tau ~|~ \tau]
%\ge
%F(q_0, T-\tau+2) - 1
%=
%\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right) \; .
%$$

@@

Combining this fact with Equations~\eqref{eqn:err-phase1} and~\eqref{eqn:err-phase2}, we have that
\begin{align*}
 \Exp \left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right]
& \ge q_0 \Exp[\tau - 1] + \Exp \left[\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right)\right] \\
& =  \Exp \left[  q_0 (\tau - 1) + \frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right)  \right] \; .
\end{align*}

%\Exp \left[~\middle|~ \tau \right]

We lower bound the last expression by considering two cases for $\tau$.
If $\tau \ge \frac{1}{2}T + 1$, then the last expression is lower bounded by
$\frac{1}{2}q_0 T = \frac{1}{2} \sqrt{T}$. If
$\tau < \frac{1}{2}T+1$, it is lower bounded by
\begin{align*}
& \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2}T}\right) \\
& = \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2q_0^2}}\right) \\
& \ge \frac{1-\frac{1}{\sqrt{2}}}{q_0}\left(1-\frac{1}{\sqrt{e}}\right) \\
& \ge \frac{1}{10} \sqrt{T} \; .
\end{align*}

### 2019-05-03 ICML-camera-ready/notions-of-linear-separability.tex
add camera-ready folder

We define two notions of linear separability for multiclass classification. The
first notion is the standard notion of linear separability used in the proof of
the mistake bound for the \textsc{Multiclass Perceptron} algorithm\citep[see e.g.][]{Crammer-Singer-2003}. The second
notion is stronger, i.e. more restrictive. %However, it is more suitable for the
%bandit setting, since it allows for a simple and efficient algorithm; see
%Section~\ref{section:algorithm-for-strongly-linearly-separable-data}.

@@

The notion of strong linear separability has appeared in the literature; see
e.g.~\citet{Chen-Chen-Zhang-Chen-Zhang-2009}. Intuitively, strong linear
separability means that, for each class $i$, the set of examples belonging to
class $i$ and the set of examples belonging to the remaining $K-1$ classes are
separated by a linear classifier $w_i$ with margin $\frac{\gamma}{2}$.

>>>>

We define two notions of linear separability for multiclass classification. The
first notion is the standard notion of linear separability used in the proof of
the mistake bound for the \textsc{Multiclass Perceptron} algorithm \citep[see e.g.][]{Crammer-Singer-2003}. The second
notion is stronger, i.e. more restrictive. %However, it is more suitable for the
%bandit setting, since it allows for a simple and efficient algorithm; see
%Section~\ref{section:algorithm-for-strongly-linearly-separable-data}.

@@

The notion of strong linear separability has appeared in the literature; see
e.g.~\citep{Chen-Chen-Zhang-Chen-Zhang-2009}. Intuitively, strong linear
separability means that, for each class $i$, the set of examples belonging to
class $i$ and the set of examples belonging to the remaining $K-1$ classes are
separated by a linear classifier $w_i$ with margin $\frac{\gamma}{2}$.

### 2019-05-03 ICML-camera-ready/related-work.tex
add camera-ready folder

The linear classification version of this problem is initially studied
by~\citet{Kakade-Shalev-Shwartz-Tewari-2008}. They proposed two computationally
inefficient algorithms that work in the weakly linearly separable setting, one
with a mistake bound of $O(K^2 d \ln(d/\gamma))$, the other with a mistake bound
of $\widetilde{O}((K^2/\gamma^2) \ln T)$. The latter result was later improved
by \citet{Daniely-Helbertal-2013}, which gives a computationally inefficient
algorithm with a mistake upper bound of $\widetilde{O}(K/\gamma^2)$. In
addition,~\citet{Kakade-Shalev-Shwartz-Tewari-2008} propose the
\textsc{Banditron} algorithm, a computationally efficient algorithm that has a
$O(T^{2/3})$ regret against the multiclass hinge loss in the general setting,
and has a $O(\sqrt{T})$ mistake bound in the $\gamma$-weakly linearly separable
setting. In contrast to mild dependencies on the time horizon for mistake bounds
of computationally inefficient algorithms, the polynomial dependence of
\textsc{Banditron}'s mistake bound on the time horizon is undesirable for
problems with a long time horizon, in the weakly linearly separable setting. One
key open question left by~\citet{Kakade-Shalev-Shwartz-Tewari-2008} is whether
one can design computationally efficient algorithms that achieve mistake bounds
that match or improve over those of inefficient algorithms. In this paper, we
take a step towards answering this question, showing that efficient algorithms
with mistake bounds quasipolynomial in $1/\gamma$ (for constant $K$) or
quasipolynomial in $K$ (for constant $\gamma$) can be obtained.

>>>>

The linear classification version of this problem is initially studied
by~\citet{Kakade-Shalev-Shwartz-Tewari-2008}. They proposed two computationally
inefficient algorithms that work in the weakly linearly separable setting, one
with a mistake bound of $O(K^2 d \ln(d/\gamma))$, the other with a mistake bound
of $\widetilde{O}((K^2/\gamma^2) \ln T)$. The latter result was later improved
by \citet{Daniely-Helbertal-2013}, which gives a computationally inefficient
algorithm with a mistake upper bound of $\widetilde{O}(K/\gamma^2)$. In
addition,~\citet{Kakade-Shalev-Shwartz-Tewari-2008} propose the
\textsc{Banditron} algorithm, a computationally efficient algorithm that has a
$O(T^{2/3})$ regret against the multiclass hinge loss in the general setting,
and has a $O(\sqrt{T})$ mistake bound in the $\gamma$-weakly linearly separable
setting. In contrast to mild dependencies on the time horizon for mistake bounds
of computationally inefficient algorithms, the polynomial dependence of
\textsc{Banditron}'s mistake bound on the time horizon is undesirable for
problems with a long time horizon, in the weakly linearly separable setting. One
key open question left by~\citet{Kakade-Shalev-Shwartz-Tewari-2008} is whether
one can design computationally efficient algorithms that achieve mistake bounds
that match or improve over those of inefficient algorithms. In this paper, we
take a step towards answering this question, showing that efficient algorithms
with mistake bounds quasipolynomial in $1/\gamma$ (for constant $K$) and
quasipolynomial in $K$ (for constant $\gamma$) can be obtained.
